Active-contours driven registration method for structure-informed segmentation of diffusion MR images

### 2015-04-06 Publications/2015-NeuroImage/05-results.tex
major update

\begin{figure*}
    \centering
    \resizebox{\textwidth}{!}{%
      \begin{tikzpicture}
        \node[inner sep=10pt, draw=black!40](fig03b) at (0,7.9)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-b}};
        \node[inner sep=10pt, draw=black!40](fig03c) at (0,0)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-c}};
        \node[circle, text=black!75] at (-9.4,10.6) {\Huge \textbf{A}};
        \node[circle, text=black!75] at (-9.4,3.5) {\Huge \textbf{B}};
      \end{tikzpicture}
    }%
  \caption{A. Visual assessment of the results on the low resolution sets:
    ``gyrus'' (top-left), ``L-shape'' (top-right), ``ball'' (bottom-left),
    and ``box'' at (bottom-right).
  In yellow color, the recovered contours after registration are represented.
  Our method showed high accuracy, as it demonstrates the almost exact location of the registered
    contours with respect to their ground truth position depicted in green.
  Partial volume effect turns segmentation of the sulci a challenging problem with voxel-wise
    clustering methods, but it is successfully segmented with our method.
  B. Quantitative evaluation of registration error in terms of average Hausdorff distance of
    surfaces at low (left) and high (right) resolutions, demonstrating that the error is
    consistently below the voxel size.
    }\label{fig:phantom}
\end{figure*}

Using a simplified subset of the experimental instrumentation described in
  \autoref{sec:experiments_evaluation}, we collected quantitative results on the digital
  phantoms.
A total of 1200 registration experiments (4 phantom types $\times$ 150 random warpings
  $\times$ 2 resolutions) were run.
Parameter settings of every processing step in the experiments are made available along with
  the source code.
For each registration, the averaged Hausdorff distance (see \autoref{sec:experiments_evaluation})
  of the inner and the outer surfaces was computed.
The results showed a consistent and high accuracy, below the image resolution.
\autoref{fig:phantom} (block C) presents the violin plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the misregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel size with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals
  (\autoref{tab:ci_phantom}).

@@

\subsection{Correction of distortions on real datasets}\label{sec:results_hcp}

\paragraph*{Assessing the segmentation model}\label{sec:res_model_and_metric} %
%
Preliminarily, we investigated the aptness of our segmentation model to play as cost function
  driving the registration process.
First, we evaluated several alternative models to segment the target bivariate images
  (\gls*{fa} and \gls*{adc}).
Based on the observed joint distribution of each (\suppl{section S4}),
  we selected the 6-regions scheme described in \autoref{sec:human_connectome}.

The distortions follow expression \eqref{eq:fieldmap} and hence, we easily reduced the
  error space to one dimension as follows.
The energy functional \eqref{eq:energy} was evaluated for the warpings given by
  a linearized space of distortions
  $\hat{U} = \epsilon \cdot U_{true} = \vec{r} + \epsilon \cdot u(\vec{r})$,
  with $\epsilon \in [-1.1, 1.1]$.
The results of this experiment are shown in \autoref{fig:energymap}.
The metric consistently displayed its minimum at $\epsilon=0.0$ (ground-truth location).

\paragraph*{Evaluation \& cross-comparison}\label{sec:res_cc_evaluation}
%
Finally, we compared the performance of \emph{regseg} and the standard \gls*{t2b}
  correction.
We computed the \gls*{swindex} \eqref{eq:swindex} of every surface after correction,
  for both \emph{regseg} and the \gls*{t2b} methods.
Finally, to statistically compare the results, we performed a one-way ANOVA test
  on the warping indices for three specific surfaces of interest
  ($\Gamma_{VdGM}$, $\Gamma_{WM}$, and $\Gamma_{pial}$).
These interfaces are critical in analyses of whole-brain tractography.
The one-way ANOVA evidenced that error distributions were significantly different,
  $p \approx 2 \times 10^{-9}$.
We also reported the 95\% CIs of the \gls*{swindex} for those surfaces.
The aggregate CI for \emph{regseg} was $[1.08 - 1.50] mm$, whereas the competing method
  yielded an aggregate CI of $[2.06 - 2.43] mm$.
The quantitative results of the ANOVA tests and the CIs are presented in \autoref{tab:results_real}.
Visual assessment of one case is shown in \autoref{fig:results_real}, along with the corresponding
  violin plots of error distributions by correction method and surface.
Extended visual assessment of the 16 cases is included in the \suppl{section S5}.

>>>>

\begin{figure*}
    \centering
    \resizebox{\textwidth}{!}{%
      \begin{tikzpicture}
        \node[inner sep=10pt, draw=black!40](fig03b) at (0,7.9)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-b}};
        \node[inner sep=10pt, draw=black!40](fig03c) at (0,0)
          {\hspace*{20pt}\includegraphics[width=\linewidth]{figures/figure03-c}};
        \node[circle, text=black!75] at (-9.4,10.6) {\Huge \textbf{A}};
        \node[circle, text=black!75] at (-9.4,3.5) {\Huge \textbf{B}};
      \end{tikzpicture}
    }%
  \caption{A. Visual assessment of the results on the low resolution sets:
    ``gyrus'' (top-left), ``L-shape'' (top-right), ``ball'' (bottom-left),
    and ``box'' at (bottom-right).
  In yellow color, the recovered contours after registration are represented.
  Our method showed high accuracy, as it demonstrates the almost exact location of the registered
    contours with respect to their ground truth position depicted in green.
  Partial volume effect turns segmentation of the sulci a challenging problem with voxel-wise
    clustering methods, but it is successfully segmented with our method.
  B. Quantitative evaluation of registration error in terms of average Hausdorff distance of
    surfaces at low (left) and high (right) resolutions, demonstrating that the error is
    consistently below the voxel size.
    }\label{fig:phantom}
\end{figure*}
A total of 1200 experiments (4 phantom types $\times$ 150 random warpings $\times$ 2 resolutions) were
  run using the workflow given in \autoref{fig:evphantoms}.
For each experiment, the misregistration error was measured using the Hausdorff distance
  (see \autoref{sec:experiments_evaluation}) between the theoretical $\gammaset_\text{true}$ and
  the estimation done by \emph{regseg} ($\hat{\gammaset}_{test}$).
The results showed a consistent and high accuracy, below the image resolution.
\autoref{fig:phantom} (block C) presents the violin plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the misregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel size with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals,
  reported in \autoref{tab:ci_phantom}.

@@

\subsection{Evaluation on real datasets and cross-comparison}\label{sec:results_hcp}
%
Finally, we compared the performance of \emph{regseg} and the standard \gls*{t2b}
  registration.
We computed the \gls*{swindex} \eqref{eq:swindex} of every surface after registration,
  for both \emph{regseg} and the \gls*{t2b} methods.
Finally, to statistically compare the results, we performed a one-way ANOVA test
  on the warping indices for three specific surfaces of interest
  ($\Gamma_{VdGM}$, $\Gamma_{WM}$, and $\Gamma_{pial}$).
These interfaces are critical in analyses of whole-brain tractography.
The one-way ANOVA evidenced that error distributions were significantly different,
  $p \approx 2 \times 10^{-9}$.
We also reported the 95\% CIs of the \gls*{swindex} for those surfaces.
The aggregate CI for \emph{regseg} was 1.08 - 1.50 [mm], whereas the \gls*{t2b} method
  yielded an aggregate CI of 2.06 - 2.43 [mm].
The quantitative results of the ANOVA tests and the CIs are presented in \autoref{tab:results_real}.
Visual assessment of the 16 cases is included in the \suppl{section S5}.
In \autoref{fig:results_real}, box A, the visual report for one subject is supplied.
Box B presents the corresponding violin plots of error distributions by registration method
  and surface.

### 2015-04-06 Publications/2015-NeuroImage/06-discussion.tex
major update

\paragraph*{Accuracy tests}
The hypothesis underpinning our research was that image registration can be reliably performed
  by searching for homogeneous regions in the target image that correspond to precise contours
  from an atlas, or extracted from other image (i.e. a different time step).
We proved that active contours without edges can be used to successfully drive a
  deformation supported by B-Spline functions using digital phantoms.
We randomly deformed 4 different phantom models mimicking three homogeneous regions,
  showing contrasts generated by a publicly available \gls*{mri} simulator.
After registration with \emph{regseg} we measured the Hausdorff distance between the
  projected contours using the ground-truth warping and our estimation.
We concluded that errors are significantly lower than voxel resolution.
We also assessed the 95\% \gls*{ci}, yielding an aggregate interval of
  $[0.64 - 0.66]mm$ in the low resolution phantoms ($2.0mm$ isotropic voxel) and
  $[0.34 - 0.38]mm$ in the high resolution phantoms ($1.0mm$ isotropic).
Therefore, we also conclude that the error is bounded and below the half of the
  voxel spacing to illustrate the accuracy of \emph{regseg}.

\paragraph*{Application on real data}
We designed \emph{regseg} as a method to correct \gls*{dmri} data for susceptibility
  distortions.
Applications of whole-brain tractography generally include the acquisition of \gls*{t1} images
  to provide with prior information about anatomy with high accuracy.
Our method solves in one single step a joint problem usually addressed in two steps:
  linear registration of \gls*{t1} and \gls*{dmri} to fuse the anatomical information on
  the diffusion data, and correction of distortion using extra acquisitions.
Moreover, our method does not require any additional acquisition over the minimal protocol
  that only includes \gls*{t1} and \gls*{dmri} images, a very common situation for
  historical datasets.
Beyond the proposed correction, potential applications of \emph{regseg} are atlas-based
  segmentation and tracking objects in time-series.

After a preliminary experiment to demonstrate the aptness of our segmentation model
  in section \autoref{sec:res_model_and_metric}, we evaluated \emph{regseg} in
  a real environment, using the experimental framework presented in \autoref{fig:evworkflows}.
We processed 16 subjects from the \gls*{hcp} database with both \emph{regseg}
  and an in-house replication of the \gls*{t2b} correction method.
\emph{Regseg} showed very high accuracy, with an aggregate 95\% \gls*{ci} of $[1.08-1.50]mm$,
  almost below the pixel size of $1.25mm$.
The misregistration error remaining after \emph{regseg} was significantly lower ($p < 0.01$) than the
  error corresponding to the \gls*{t2b} correction, using the one-way ANOVA test
  (\autoref{tab:results_real}).
Visual inspection of all the results (\suppl{section S5}), and the violin plots included in
  \autoref{fig:results_real}, confirmed this intuition.
Even though we carefully configured the \gls*{t2b} method by using the same algorithm and
  same settings from a widely used software, cross-comparison experiments are prone to
  the so-called \emph{instrumentation bias} \citep{tustison_instrumentation_2013}.
Therefore, these results do not show that the \emph{regseg} software \emph{is} better than the
  \gls*{t2b} method, but they are communicated to proof that \emph{regseg} is a solid option to
  correct \gls*{dmri} data for distortions, and any other potential application.

\paragraph*{Prospects}
Future improvements of \emph{regseg} possibly will follow a similar line to the evolution of 
  existing methodologies.
Incorporating knowledge about the distortion by initializing our method with the theory-based
  correction using fieldmaps may be the most cost-effective.
The main pitfall using fieldmap correction is the challenging problem of unwrapping the
  range of the phase map, as it is clipped in the range $(-\pi, \pi]$.
The most extended solution to this is \emph{prelude} \citep{jenkinson_fast_2003}, and some
  alternatives have been proposed \citep{zhou_reliable_2009,daga_susceptibility_2014} to
  circumvent the inherent unreliability of the method.
The \emph{regseg} algorithm would naturally correct misregistered areas due to errors in
  phase unwrapping, and underestimations introduced by necessary smoothing
  applied to clean up field maps.
This integration has been recently applied in the \gls*{t2b} approach
  by \cite{irfanoglu_susceptibility_2011}, also investigating the usefulness of
  nonregular B-Spline grids.
This is also a potential extension to our deformation model.

Finally, we are currently studying substitute features to build the energy functional on,
  enabling to perform \emph{regseg} directly on the raw \gls*{dmri} data.

>>>>

\paragraph*{Accuracy tests}
The hypothesis underpinning our research was that image registration can be reliably performed
  by searching for homogeneous regions in the target image that correspond to precise contours
  from an atlas, or extracted from other image (i.e. a different time step).
We proved that active contours without edges can be used to successfully drive a
  deformation supported by B-Spline basis functions using digital phantoms.
We randomly deformed 4 different phantom models mimicking three homogeneous regions,
  and use them to simulate \gls*{t1} and \gls*{t2} images at two resolution levels.
After registration with \emph{regseg} we measured the Hausdorff distance between the
  projected contours using the ground-truth warping and our estimation.
We concluded that errors are significantly lower than voxel resolution.
We also assessed the 95\% \gls*{ci}, yielding an aggregate interval of
  0.64 - 0.66 [mm] in the low resolution phantoms (2.0mm isotropic voxel) and
  0.34 - 0.38 [mm] in the high resolution phantoms (1.0mm isotropic).
Therefore, we also conclude that the error is bounded above by the half of the
  voxel spacing.

\paragraph*{Application on real data}
We designed \emph{regseg} as a method to segment \gls*{dmri} data exploiting the
  anatomical information extracted in the corresponding \gls*{t1} image of the subject.
Applications of whole-brain tractography \citep{smith_tractbased_2006,craddock_imaging_2013}
  usually solve the problem in a two-step approach.
First, images are corrected for nonlinear distortions using auxiliary acquisitions
  such as fieldmaps \citep{jezzard_correction_1995}, \gls*{t2} images \citep{kybic_fast_2003},
  etc.
Second, the segmentation is projected from a reference \gls*{t1} image using linear
  registration \citep{greve_accurate_2009}.
\emph{Regseg} solves this joint problem in one single step, and it does not require any additional
  acquisition over the minimal protocol including \gls*{t1} and \gls*{dmri} images only.
This situation is commonly found in historical datasets.
Beyond the proposed application in \gls*{dmri} data, other potential uses of \emph{regseg} are
  atlas-based segmentation and tracking objects in time-series.

We evaluated \emph{regseg} in a real environment, using the experimental framework presented
  in \autoref{fig:evworkflows}.
We processed 16 subjects from the \gls*{hcp} database with both \emph{regseg}
  and an in-house replication of the \gls*{t2b} correction method.
\emph{Regseg} showed very high accuracy, with an aggregate 95\% \gls*{ci} of 0.56-0.66 [mm],
  bellow the pixel size of 1.25mm.
The misregistration error remaining after \emph{regseg} was significantly lower ($p < 0.01$) than the
  error corresponding to the \gls*{t2b} correction, using the one-way ANOVA test
  (\autoref{tab:results_real}).
Visual inspection of all the results (\suppl{section S5}), and the violin plots included in
  \autoref{fig:results_real}, confirmed this intuition.
Even though we carefully configured the \gls*{t2b} method by using the same algorithm and
  same settings of a widely used software, cross-comparison experiments are prone to
  the so-called \emph{instrumentation bias} \citep{tustison_instrumentation_2013}.
Therefore, these results do not show that the \emph{regseg} software \emph{is} better than the
  \gls*{t2b} method, but they are communicated to proof that \emph{regseg} is a solid option
  in the proposed application.
Additionally, the \gls*{t2b} may introduce an additional (and small) error in the necessary
  registration of \gls*{t2} to \gls*{t1} space.

% \paragraph*{Prospects}
% First extensions of this work will study more appropriate features to build the energy functional
%   on, enabling to perform \emph{regseg} directly on the raw \gls*{dmri} data.
% A second outlook covers incorporating knowledge about the distortion by initializing our method
%   with the theory-based displacement field that can be estimated with fieldmaps.

### 2015-04-06 Publications/2015-NeuroImage/07-conclusion.tex
major update

\emph{Regseg} is a variational framework to simultaneously segment and
  register 3D \gls*{dmri} data of the human brain, using within-subject
  anatomical information as reference.
The registration method segments the target multivariate image in several competing regions
  defined explicitly by their limiting surfaces.
The surfaces are active, and evolve on a free-deformation field supported by B-Splines.
A descent optimization strategy is guided by shape gradients computed on the current partition
  of the target image.
\emph{Regseg} uses active contours without edges, and hence looks for
  homogeneous regions within the image.
We tested \emph{regseg} on digital phantoms simulating \gls*{t1} and \gls*{t2} \gls*{mri},
	warped with smooth random deformations.
The resulting average misplacement of the contours was significantly lower than the
  image resolution of the phantoms.

We propose \emph{regseg} to correct \gls*{dmri} datasets for the typically present distortions
  caused by inhomogeneities of the main magnetic field.
Visual assessment of \emph{regseg} and cross-comparison with a widely-used competing
  technique demonstrated its accuracy.
Moreover, \emph{regseg} does not require additional images to the minimal acquisition protocol
  that includes only \gls*{t1} and \gls*{dmri}.

\section*{Availability and reproducibility statement}
\label{sec:availability}
We considered the reproducibility of our results a design requirement.
Therefore, we used real data fetched from a publicly available repository
  (the Human Connectome Project \citep{essen_human_2012}) and all the software
  involved in this paper is also publicly released.
The \emph{regseg} algorithm is implemented in C++, on top of ITK-4.6
  (Insight Registration and Segmentation Toolkit, \url{http://www.itk.org}).
The evaluation instruments (\autoref{fig:evworkflows}), are implemented in Python using
  \emph{nipype} \citep{gorgolewski_nipype_2011} to assess their reproducibility.
Phantom data are automatically generated by the evaluation framework.
All research elements (data, source code, figures, manuscript sources, etc.) in this paper
  are made publicly available under a unique digital object identifier in a
  package \citep{esteban_acweregistration_2015}.

>>>>

\emph{Regseg} is a variational framework to simultaneously segment and
  register 3D \gls*{dmri} data of the human brain, using within-subject
  anatomical information as reference.
The registration method segments the target multivariate image in several competing regions
  defined explicitly by their limiting surfaces.
The surfaces are active, and evolve on a free-form deformation field supported by B-Splines.
A descent optimization strategy is guided by shape gradients computed on the current partition
  of the target image.
\emph{Regseg} uses active contours without edges, and hence looks for
  homogeneous regions within the image.
We tested \emph{regseg} on digital phantoms simulating \gls*{t1} and \gls*{t2} \gls*{mri},
	warped with smooth random deformations.
The resulting misregistration of the contours was significantly lower than the image resolution
  of the phantoms.

We propose \emph{regseg} to simultaneously segment \gls*{dmri} data and register them to
  their corresponding \gls*{t1} image of the same subject.
Visual assessment of \emph{regseg} and cross-comparison with a widely-used technique demonstrated 
  its accuracy.
Moreover, \emph{regseg} does not require additional images to the minimal acquisition protocol
  that includes only \gls*{t1} and \gls*{dmri}.

\section*{Availability and reproducibility statement}
\label{sec:availability}
We considered the reproducibility of our results a design requirement.
Therefore, we used real data fetched from a publicly available repository
  (the Human Connectome Project \citep{essen_human_2012}) and all the software
  involved in this paper is also publicly released.
\emph{Regseg} is implemented on top of ITK-4.6 (Insight Registration and 
  Segmentation Toolkit, \url{http://www.itk.org}).
The evaluation instruments (\autoref{fig:evworkflows}), are implemented using
  \emph{nipype} \citep{gorgolewski_nipype_2011} to assess their reproducibility.
All research elements (data, source code, figures, manuscript sources, etc.) in this paper
  are made publicly available under a unique package \citep{esteban_acweregistration_2015}.
